Owlifier: Creating OWL-DL Ontologies from Simple Spreadsheet-Based Knowledge Descriptions\tnoteref{t1}

### 2008-12-03 docs/pubs/ISEI6/main.tex

\myblock{Sufficient Blocks.} Sufficient blocks are similar to synonym
blocks in that they state equivalences between classes. However,
equivalence relationships are defined in sufficient blocks through
property restrictions (i.e., via anonymous classes). We consider two
main types of sufficient blocks. Rows containing sufficient blocks
take one of the following forms.
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|l|l|l|l|l|}\hline 
    \textsf{sufficient} & \textsf{min}, \textsf{max}, or 
    \textsf{exact} & $m$ & $p$ & $c_1$ & $c_2$ & \dots & $c_n$\\ \hline
  \end{tabular} \hfill $(n \ge 2)$
\item[]
  \begin{tabular}{|l|l|l|l|l|l|l|}\hline 
    \textsf{sufficient} & \textsf{only} & $p$ & $c_1$ & $c_2$ & \dots & 
    \{\textsf{not}\} $c_n$\\ \hline
  \end{tabular} \hfill $(n \ge 2)$
\end{itemize}
The first form states that individuals $p$-related to at least, at
most, or exactly $m$ individuals of class $c_{i+1}$ are by definition
individuals of the class $c_i$, for $1 \le i < n$. Each class $c_i$ of
a sufficient block of this form induces either the description logic
axiom $c_i \equiv \; (\le m) \; p . c_{i+1}$, $c_i \equiv \; (\ge m)
\; p . c_{i+1}$, or $c_i \equiv \; (= m) \; p . c_{i+1}$ depending on
whether the qualifier \textsf{min}, \textsf{max}, or \textsf{exact} is
used, respectively.

The second form states that individuals $p$-related to only
individuals of class $c_{i+1}$ are by definition individuals of the
class $c_i$, for $1 \le i < n$. For each $c_i$ a sufficient block of
this form induces the description logic axiom $c_i \equiv \forall p
. c_{i+1}$. The complement of the last class in a sequence may also be
used within these blocks via the expression \textsf{not
  $c_n$}. Sufficient blocks of this form induce description logic
axioms $c_{n-1} \equiv \forall p . \neg c_n$ stating that individuals
that are $p$-related to only individuals that are not of class $c_n$
(i.e., the complement of $c_n$) are by definition individuals of class
$c_{n-1}$.  Multiple sufficient blocks for a class are treated as a
single description logic axiom through intersection. For example, the
blocks
\begin{tabbing}
  ~~\textsf{sufficient} \= \textsf{only} \= $m$ \= $c_1$ \= $c_2$ \= \kill
  ~~\textsf{sufficient} \> \textsf{min} \> $m$ \> $p$ \> $c_1$ \> $c_2$ \\ 
  ~~\textsf{sufficient} \> \textsf{max} \> $m$ \> $p$ \> $c_1$ \> $c_2$ \\ 
  ~~\textsf{sufficient} \> \textsf{only} \> $p$ \> $c_1$ \> $c_2$
\end{tabbing}
result in the single axiom $c_1 \equiv \; (\le m \; p .  c_2) \sqcap
(\ge m \; p . c_2) \sqcap (\forall p.c_2)$. The following blocks are
simple examples of sufficient conditions.
\begin{tabbing}
  ~~\textsf{sufficient} \textsf{exact} \= \kill
  ~~\textsf{sufficient} \textsf{min} \> 1 \textsf{hasPart} 
  \textsf{Mammal} \textsf{Hair} \\
  ~~\textsf{sufficient} \textsf{exact} \> 2  \textsf{hasPart} 
  \textsf{Mammal} \textsf{MammalianEar} \\
  ~~\textsf{sufficient} \textsf{only} \> \textsf{feedsYoungWith} 
  \textsf{Mammal} \textsf{MammaryGland}\\
  ~~\textsf{sufficient} \textsf{exact} \> 3 \textsf{hasPart} 
  \textsf{MammalianEar} \textsf{MiddleEarBone}
\end{tabbing}
The first three blocks state that individuals that have Hair as a
part, that have exactly two Mammalian Ears as parts, and that feed
their young via Mammary Glands are Mammals. The last block states
individuals that have exactly three Middle Ear Bones are Mammalian
Ears.

@@

@@

2. disjoint classes are computed by default

3. property closures

@@

future work: the ongoing extensions described above, support seamless
conversion into and out of owlifier, e.g., if someone extends owlifier
ontology, re-apply extensions after new owlifier changes. Also,
support translation to oboe.

>>>>

\myblock{Sufficient Blocks.} Sufficient blocks are similar to synonym
blocks in that they state equivalences between classes. A sufficient
block serves as a modifier to entity blocks, relationship blocks
(including transitive blocks), and cardinality blocks, modifying the
associated description logic axioms by using equivalence ($\equiv$) in
place of sublass ($\sqsubseteq$). For example the block
\begin{tabbing}
  ~~\textsf{sufficient} \textsf{relationship} \textsf{hasPart} \textsf{Mammal} \textsf{Hair} 
\end{tabbing}
induces the description logic axiom $\textsf{Mammal} \equiv \exists
\textsf{hasPart}.\textsf{Hair}$ stating that \emph{any} individual
that has Hair as a part is a Mammal. Additionally, we allow
relationship blocks within a sufficient block to be extended with
\textsf{not} to state that absense of the property is a defining
characteristic of the class. For example the following blocks
\begin{tabbing}
  ~~\textsf{sufficient} \textsf{entity} \textsf{Mammal} \textsf{Eutheria} \\
  ~~\textsf{sufficient} \textsf{relationship not} \textsf{hasPart} \textsf{Eutheria} \textsf{EpipubicBone} 
\end{tabbing}
induces the axiom $\textsf{Eutheria} \equiv \textsf{Mammal} \sqcap
\neg \exists \textsf{hasPart} . \textsf{EpipubicBone}$ stating that
any Mammal that does not have an Epipubic Bone as a part is a
Eutheria. Note also that multiple sufficient blocks for a particular
class result in a single axiom in which each constraint is combined
via conjunction ($\sqcap$).

@@

In this section, we briefly describe some of the additional features
of \owlifier, specifically focusing on the use of disjoint classes,
property closures, \owlifier\ reasoning, and additional block syntax.

\subsection{Disjoint Class Inference}

OWL is based on the open world assumption, which can lead to a number
of ontology development ``pitfalls'' for those new to the language
\cite{smith04:_owl_web_ontol_languag_guide,rector04:_owl_pizzas}. One
example is in the creation of disjoint classes (another concerns
property closures, described below). In particular, unless explicitly
stated, distinct classes within an OWL-DL ontology are never assumed
to be disjoint. However, in many ontologies a large number of classes
are typically defined as being disjoint (e.g., sibling classes), and
stating these disjoint constraints is often time-consuming since each
pair of classes must be given an explicit disjoint assertion. Editors
such as Protege \cite{knublauch04:_editin_descr_logic_ontol_with}
provide shortcuts via the user interface to create specific sets of
disjoint assertions, e.g., by allowing a user to define all children
of a particular class as disjoint. In general, however, many users
expect such classes to be disjoint by default
\cite{rector04:_owl_pizzas} and this expectation often leads to
modeling errors.

Alternatively, the default assumption in \owlifier\ is that distinct
classes are disjoint. Specifically, as an \owlifier\ table is
converted to an OWL-DL ontology, the system analyzes the class
hierarchy structure and identifies pairs of classes that are: (1) not
related via subclass relations (either direct or indirect subclasses);
(2) not defined as synonyms; and (3) not explicitly defined to overlap
via an overlap block. Each such pair of classes is then asserted by
\owlifier\ in the resulting ontology as being disjoint. As described
in \cite{rector04:_owl_pizzas}, undeclared disjoint classes are a
common problem in ontology development using OWL-DL and often limit
the utility of reasoning systems (by limiting the inferences that can
be obtained). The approach employed in \owlifier\ for handling
disjoint classes makes the common expectations of users the default
case, which in general should lead to a more intuitive ontology
editing environment and an overall fewer number of modeling mistakes.

\subsection{Object Property Closure}

Another common mistake for new users of OWL-DL concerns the user of
universal rather than existential property restrictions
\cite{rector04:_owl_pizzas}. In particular, an expression such as
$\texttt{Body} \sqsubseteq \forall \textsf{hasPart}$

\subsection{Reasoning in \owlifier}

\subsection{Sparse Blocks}

@@

@@

future work: the ongoing extensions described above, support seamless
conversion into and out of owlifier, e.g., if someone extends owlifier
ontology, re-apply extensions after new owlifier changes. Also,
support translation to oboe. Additional testing and evaulation.
